### 2021-11-23 text/outline/draft.tex
finish draft of simulations

\begin{abstract}
CRISPR genome engineering and single-cell sequencing have transformed biological discovery. Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression and illuminating regulatory networks underlying diseases. Despite their promise, single-cell CRISPR screens present substantial statistical challenges. We demonstrate on real data that a standard method for estimation and inference in single-cell CRISPR screens — “thresholded regression” — exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic tuning parameter. We recover these phenomena in precise theoretical terms in an idealized Gaussian setting. Next, we introduce GLM-EIV (``exponential family errors-in-variables model''), a new method for single-cell CRISPR screen analysis. GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed, overcoming limitations of thresholded regression. We develop a computational infrastructure to deploy GLM-EIV across hundreds or thousands of processors on clouds (e.g., Microsoft Azure) and high-performance clusters. Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets, yielding new biological insights.
\end{abstract}
\tableofcontents
\newpage
\section{Introduction}
CRISPR is a genome engineering tool that has enabled scientists to precisely edit human and nonhuman genomes, opening the door to new medical therapies \cite{Rothgangl2021,Musunuru2021} and transforming basic biology research \cite{Przybyla2021}. Recently, scientists have paired CRISPR genome engineering with single-cell sequencing \cite{Dixit2016,Datlinger2017}. The resulting assays, known as a ``single-cell CRISPR screens,'' link genetic perturbations in individual cells to changes in gene expression, illuminating regulatory networks underlying human diseases and other traits \cite{Morris2021a}.

@@

Our work makes two main contributions. First, we conduct a careful study of the thresholding method. Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in an idealized Gaussian setting. Second, we introduce a new method, GLM-EIV (generalized linear model with errors-in-variables), for single-cell CRISPR screen analysis. GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed. GLM-EIV thereby implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic. Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.

We implement several statistical accelerations (that likely are of independent utility) to bring the cost of GLM-EIV down to within an order of magnitude of the thresholding method. Finally, we develop a computational infrastructure to deploy GLM-EIV at-scale across hundreds or thousands of processors on clouds (e.g., Microsoft Azure) and high-performance clusters. Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, high multiplicity-of-infection single-cell CRISPR screen datasets, yielding new biological and statistical insights.

@@

Motivated by the challenges of single-cell data, several authors recently have extended statistical models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broader class of exponential family distributions. For example, Lin, Lei, and Roeder \cite{Lin2021} developed eSVD, an extension of SVD to exponential family and curved Gaussian responses. Unlike SVD, eSVD models the relationship between the mean and variance of a gene's expression level, a phenomenon induced by the countedness of single-cell data \cite{Lause2021}.
Similarly, Townes et al.\ \cite{Townes2019} proposed GLM-PCA, a generalization of PCA that directly models negative binomially-distributed gene expression counts while regressing out technical factors. We see our work as a continuation of this broad effort to ``port'' common statistical methods and models to single-cell count data. Our focus, however, is on regression rather than dimension reduction: we extend the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed.

@@

The first step (Algorithm \ref{algo:pilot_estimates}, lines 2-7) is to obtain good parameter estimates for $[\beta^m_0, \gamma_m]^T$ and $[\beta^g_0, \gamma_g]^T$ using regression. Recall that the underlying gene expression parameter vector $\beta_m$ is $\beta_m = [\beta^m_0, \beta^m_1, \gamma_m]^T \in \R^d$, where $\beta^m_0$ is the intercept, $\beta^m_1$ is the effect of the perturbation, and $\gamma_m^T$ is the effect of the technical factors. To produce estimates $[\beta^m_0]^\textrm{pilot}$ and $[\gamma_m^T]^\textrm{pilot}$, we regress the gene expressions $m$ onto the technical factors $X$. The intuition for this procedure is as follows: the probability of perturbation $\pi$ is very small. Therefore, the true log likelihood is approximately equal to the log likelihood that results from omitting $p_i$ from the model:
\begin{multline*}
\sum_{i=1}^n f_m(m_i; \eta^m_i) = \sum_{i=1}^n f_m(m_i; h_m(\beta_0 + \beta_1 p_i + \gamma^T z_i + o^m_i)) \\ = \underbrace{\sum_{p_i =1}^n f_m(m_i; h_m(\beta_0 + \beta_1 + \gamma^T z_i + o^m_i))}_\textrm{few terms}  + \underbrace{\sum_{p_i = 0}^n f_m(m_i; h_m(\beta_0 + \gamma^T z_i + o^m_i))}_\textrm{many terms} \\ \approx \sum_{i = 1}^n f_m(m_i; h_m(\beta_0 + \gamma^T z_i + o^m_i)).
\end{multline*}
We similarly can obtain pilot estimates $[\beta^g_0]^\textrm{pilot}$ and $[\gamma^T_g]^\textrm{pilot}$ by regressing the gRNA counts $g$ onto the technical factors $X$. We extract the fitted values (on the scale of the linear component) for use in a subsequent step: $$\hat{f}^k_i = [\beta^k_0]^\textrm{pilot} + \langle [\gamma^T_k]^\textrm{pilot}, z_i \rangle + o^k_i,$$
for $k \in \{m,g\}.$

@@

@@

\section{Additional simulation results}

>>>>

\begin{abstract}
CRISPR genome engineering and single-cell sequencing have transformed biological discovery. Single-cell CRISPR screens unite these two technologies, linking genetic perturbations in individual cells to changes in gene expression and illuminating regulatory networks underlying diseases. Despite their promise, single-cell CRISPR screens present substantial statistical challenges. We demonstrate on real data that a standard method for estimation and inference in single-cell CRISPR screens — “thresholded regression” — exhibits attenuation bias and a bias-variance tradeoff as a function of an intrinsic tuning parameter. We recover these phenomena in precise theoretical terms in an idealized Gaussian setting. Next, we introduce GLM-EIV (``GLM-based errors-in-variables''), a new method for single-cell CRISPR screen analysis. GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed, overcoming limitations of thresholded regression. We develop a computational infrastructure to deploy GLM-EIV across tens or hundreds of nodes on clouds (e.g., Microsoft Azure) and high-performance clusters. Leveraging this infrastructure, we apply GLM-EIV to analyze two recent, large-scale, single-cell CRISPR screen datasets, yielding new biological insights.
\end{abstract}
\tableofcontents
\newpage
\section{Introduction}
CRISPR is a genome engineering tool that has enabled scientists to precisely edit human and nonhuman genomes, opening the door to new medical therapies \cite{Rothgangl2021,Musunuru2021} and transforming basic biology research \cite{Przybyla2021}. Recently, scientists have paired CRISPR genome engineering with single-cell sequencing \cite{Dixit2016,Datlinger2017}. The resulting assays, known as a ``single-cell CRISPR screens,'' link genetic perturbations in individual cells to changes in gene expression, illuminating regulatory networks underlying human diseases and other traits \cite{Morris2021a}.

@@

Our work makes two main contributions. First, we conduct a detailed study of the thresholding method. Notably, we demonstrate on real data that the thresholding method exhibits attenuation bias and a bias-variance tradeoff as a function of the selected threshold, and we recover these phenomena in precise mathematical terms in an idealized Gaussian setting. Second, we introduce a new method, GLM-EIV (``GLM-based errors-in-variables''), for single-cell CRISPR screen analysis. GLM-EIV generalizes the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed. GLM-EIV thereby implicitly estimates the probability that each cell was perturbed, obviating the need to explicitly impute perturbation assignments via thresholding or another heuristic. Theoretical analyses and simulation studies indicate that GLM-EIV outperforms the thresholding method in large regions of the parameter space.

We implement several statistical accelerations (that likely are of independent utility) to bring the cost of GLM-EIV down to within about an order of magnitude of the thresholding method. Finally, we develop a Docker-containerized application to deploy GLM-EIV at-scale across tens or hundreds of nodes on clouds (e.g., Microsoft Azure) and high-performance clusters. Leveraging this application, we apply GLM-EIV to analyze two recent, large-scale, high multiplicity-of-infection single-cell CRISPR screen datasets, yielding new biological and statistical insights.

@@

Motivated by the challenges of single-cell data, several authors recently have extended statistical models that (implicitly or explicitly) assume Gaussianity and homoscedasticity to a broader class of exponential family distributions. For example, Lin, Lei, and Roeder \cite{Lin2021} developed eSVD, an extension of SVD to exponential family and curved Gaussian responses. Unlike SVD, eSVD models the relationship between the mean and variance of a gene's expression level, a phenomenon induced by the countedness of single-cell data \cite{Lause2021}.
Similarly, Townes et al.\ \cite{Townes2019} proposed GLM-PCA, a generalization of PCA that directly models negative binomially-distributed gene expression counts while controlling for technical factors. We see our work as a continuation of this broad effort to ``port'' common statistical methods and models to single-cell count data. Our focus, however, is on regression rather than dimension reduction: we extend the classical errors-in-variables model to response distributions and sources of measurement error that are exponential family-distributed.

@@

The first step (Algorithm \ref{algo:pilot_estimates}, lines 2-7) is to obtain good parameter estimates for $[\beta^m_0, \gamma_m]^T$ and $[\beta^g_0, \gamma_g]^T$ using regression. Recall that the underlying gene expression parameter vector $\beta_m$ is $\beta_m = [\beta^m_0, \beta^m_1, \gamma_m]^T \in \R^d$, where $\beta^m_0$ is the intercept, $\beta^m_1$ is the effect of the perturbation, and $\gamma_m^T$ is the effect of the technical factors. To produce estimates $[\beta^m_0]^\textrm{pilot}$ and $[\gamma_m^T]^\textrm{pilot}$, we regress the gene expressions $m$ onto the technical factors $X$. The intuition for this procedure is as follows: the probability of perturbation $\pi$ is very small. Therefore, the true log likelihood is approximately equal to the log likelihood that results from omitting $p_i$ from the model:
\begin{multline*}
\sum_{i=1}^n f_m(m_i; \eta^m_i) = \sum_{i=1}^n f_m(m_i; h_m(\beta_0 + \beta_1 p_i + \gamma^T z_i + o^m_i)) \\ = \underbrace{\sum_{i : p_i =1} f_m(m_i; h_m(\beta_0 + \beta_1 + \gamma^T z_i + o^m_i))}_\textrm{few terms}  + \underbrace{\sum_{i : p_i = 0} f_m(m_i; h_m(\beta_0 + \gamma^T z_i + o^m_i))}_\textrm{many terms} \\ \approx \sum_{i = 1}^n f_m(m_i; h_m(\beta_0 + \gamma^T z_i + o^m_i)).
\end{multline*}
We similarly can obtain pilot estimates $[\beta^g_0]^\textrm{pilot}$ and $[\gamma^T_g]^\textrm{pilot}$ by regressing the gRNA counts $g$ onto the technical factors $X$. We extract the fitted values (on the scale of the linear component) for use in a subsequent step: $$\hat{f}^k_i = [\beta^k_0]^\textrm{pilot} + \langle [\gamma^T_k]^\textrm{pilot}, z_i \rangle + o^k_i,$$
for $k \in \{m,g\}.$

@@

We conducted a simulation study to compare the empirical performance of GLM-EIV to that of the thresholding method. We generated data on $n = 150,000$ cells from the GLM-EIV model (Section \ref{sec:model}), setting the the target of inference $\beta^m_1$ to $\log(0.25)$ and the probability of perturbation $\pi$ to $0.02$. $\beta^m_1 = \log(0.25)$ represents a decrease in gene expression by a factor of $4$, which is a fairly large effect size on the order of what we might observe for a positive control pair. We included ``sequencing batch'' (modeled as a Bernoulli-distributed variable) as a covariate and sequencing depth (modeled as a Poisson-distributed variable) as an offset. We varied the log-fold change in gRNA expression, $\beta^g_1,$ over a grid on the interval $[\log(1), \log(4)];$ $\beta^g_1$ controls problem difficulty, with higher values corresponding to easier problems. Finally, we generated the gene expression and gRNA count data from two response distributions: Poisson and negative binomial (size parameter fixed at $\theta = 20$ for the latter). For each parameter setting (defined by a $\beta^g_1$-distribution pair), we synthesized $n_\textrm{sim} = 500$ i.i.d.\ datasets. Section \ref{sec:extra_sims} presents additional simulation results on Gaussian response distributions.

\begin{figure}
	\centering
	\includegraphics[width=0.9\linewidth]{../../figures/main_text_sim/plot}
	\caption{\textbf{Simulation study}. Columns correspond to distributions (Poisson, NB with known $\theta$, NB with estimated $\theta$), and rows correspond to metrics (bias, MSE, coverage, CI width, and time). Methods are shown in different colors; GLM-EIV (red) is masked by accelerated GLM-EIV (blue) in several panels. GLM-EIV demonstrated superior statistical performance to the thresholding method on all metrics (rows 1-4). Accelerated GLM-EIV had substantially lower computational cost than ``vanilla'' GLM-EIV (bottom row) despite demonstrating identical statistical performance (rows 1-4).}
	\label{main_text_sim}
\end{figure}

We applied three methods to the simulated data: ``vanilla'' GLM-EIV (Algorithm \ref{algo:em_full}), accelerated GLM-EIV (Algorithm \ref{algo:pilot_estimates}), and thresholded regression. We used the Bayes-optimal decision boundary for classification as the threshold for the thresholding method. We ran all methods on the negative binomial data twice: once treating the size parameter $\theta$ as a known constant and once treating $\theta$ as unknown. In the latter case we used the \texttt{glm.nb} function from the \texttt{MASS} package to estimate $\theta$ \cite{Ripley2013}. We display the results of the simulation study in Figure \ref{main_text_sim}. Columns correspond to distributions (i.e., Poisson, negative binomial with known $\theta$, and negative binomial with unknown $\theta$), rows correspond to performance metrics (i.e., bias, mean squared error, CI coverage rate (nominal rate $95\%$), CI width, and method execution time). The problem difficulty parameter $\beta^g_1$ is plotted on the horizontal axis, and the methods are depicted in different colors (GLM-EIV = red, accelerated GLM-EIV = blue, and thresholding = purple; GLM-EIV masked by accelerated GLM-EIV in several panels).

First, we observed that GLM-EIV dominated thresholded regression on all statistical metrics: GLM-EIV exhibited lower bias (row 1) and mean squared error (row 2) than thresholded regression; additionally, GLM-EIV had superior confidence interval coverage (row 3) despite having produced generally narrower confidence intervals (row 4). Intuitively, GLM-EIV outperformed the thresholding method because (i) GLM-EIV leveraged information from \textit{both} modalities (rather than the gRNA modality alone) to assign perturbation identities to cells, and (ii) GLM-EIV produced soft rather than hard assignments, capturing the inherent uncertainty in whether a perturbation occurred. Next, we found that accelerated GLM-EIV performed as well as ``vanilla'' GLM-EIV on the statistical metrics (rows 1-4) while substantially reducing computational cost (bottom row). In fact, the execution time of accelerated GLM-EIV was approximately within an order of magnitude of that of the thresholding method (bottom row).

Interestingly, thresholded regression exhibited better confidence interval coverage under estimated $\theta$ than under known $\theta$ (row 3). Estimating $\theta$ leads to slight inflation bias (i.e., overestimating the true effect size), whereas, as we showed previously, thresholding leads to attenuation bias (i.e., underestimating the true effect size). These phenomena partially cancel, yielding less biased estimates. We note that GLM-EIV and the thresholding method in principal are compatible with \textit{any} $\theta$ estimation procedure, including those based on more sophisticated techniques, such as regularization \cite{Hafemeister2019}. We defer rigorous investigation of the impact of different $\theta$ estimation strategies on these methods to future work.

@@

\section{Additional simulation study}\label{sec:extra_sims}
We ran an additional simulation study in which we modeled the gene and gRNA expressions using a Gaussian distribution with identity link. We generated data on $n = 150,000$ cells, fixing the target of inference $\beta^m_1$ to $-4$ and the probability of perturbation $\pi$ to $0.05$. We included ``sequencing batch'' (modeled as a Bernoulli-distributed variable) and ``sequencing depth'' (modeled as a Poisson-distributed variable) as covariates in the model. We did not include sequencing depth as an offset because use of the identity link renders offsets meaningless. We varied $\beta^g_1$ over a grid on the interval $[0,7].$ We generated $n_\textrm{sim} = 1,000$ synthetic datasets for each value of $\beta^g_1$. We applied accelerated GLM-EIV and thresholded regression to the simulated data. We assessed these methods on the metrics of bias, mean squared error, confidence interval coverage rate, and confidence interval width. We found that accelerated GLM-EIV outperformed the thresholding method: the former method exhibited smaller bias, smaller mean squared error, higher confidence interval coverage rate, and smaller confidence interval width than the latter method (Figure \ref{fig:gaussian_sim}).

\begin{figure}
	\centering
	\includegraphics[width=1.0\linewidth]{../../figures/supplement_sim/gaussian}
	\caption{Additional simulation results on Gaussian data. GLM-EIV (accelerated) outperformed the thresholding method on bias, mean squared error, confidence interval coverage rate, and confidence interval width metrics.}\label{fig:gaussian_sim}
\end{figure}
